---
title: "Mitochondrial ribosome profiling analysis"
author: "Sophia H. Li & Raphael J. Morscher"
date: "09/18/2017"
output:
  html_document:
    toc: true
    theme: united

---


This R markdown file contains functions used to generate ribosome profiling graphs for Morscher et al. Nature 2017. 

# Python code to generate codon count table for subseqeunt analysis in R

 * Command line in terminal - 1: Generate count table
python BAM_to_count_table.py --bam xxxx.bam --gff chrM_fixed.gff3 --fasta 170822_chrM.fa -o xxxx_codon_count.txt --add_three --offset 14 --min_length 25 --max_length 40 --threeprime

 * Command line in terminal - 2: Collapse all tables from each sample into one for analysis in R
python Collapse_codon_count_tables.py

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(RColorBrewer)
library(knitr)
library(ggrepel)
library(ineq)
opts_chunk$set(cache=TRUE, echo=TRUE, results="asis", message=FALSE)
```

# Analysis of HCT116 knockout clones
# Load and normalize data

```{r load_data}

raw_data <- read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/HCT116_WT_TO1_PatFib_all_codon_count.txt")


# Load the mapping where we can turn IDs into names that are easily identifiable 
mito_info <-read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/Mitochondrial_Genome_Transcript_GeneInformation.txt") 


condition1 <- "HCT116_WT"
condition2 <- "HCT116MTO1"
condition_to_analyze <- c(condition1,condition2)

raw_data <- raw_data %>% 
    left_join(.,mito_info,by = c("gene_id"="UCSC_id")) %>% select(gene_name = AssociatedGeneName,GeneType,sample,codon_seq:position_3_count,Complex)  %>% 
  separate(sample, into= c("condition","replicate"),remove = FALSE,sep ="-") %>% 
  filter(condition %in% condition_to_analyze)

# Load the mitochondria codon table 
mito_aminoacid_codon <- read_csv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/AA_Codon_HumMito.csv")

mito_pallete = c("black","red")

```


```{r normalizing_counts}

# Normalize counts in each sample to RPM
mito_data <- raw_data %>% select(-contains("position")) %>% group_by(sample,condition,replicate) %>%  
  mutate(RPM = codon_count_sum/sum(codon_count_sum) * 10^6 + 1) %>% 
  group_by(sample,gene_name) %>%  
  mutate(RPM_normgene = RPM/sum(RPM),
         RPM_cumsum = cumsum(RPM),
         RPM_cumsum_normgene = RPM_cumsum/sum(RPM))  %>% 
  mutate(condition = factor(condition,levels=c(condition1, condition2))) %>% 
  mutate_at(vars(codon_seq),funs(toupper))

# Summarise the replicates and generate cumsum. 
mito_data_samplemean <- mito_data %>% group_by(condition,codon_index,gene_name,codon_seq) %>% 
    summarise(RPM_samplemean = mean(RPM),
              RPM_samplemean_se = sd(RPM)/sqrt(n()),
              RPM_normgene_samplemean = mean(RPM_normgene),
              RPM_normgene_samplemean_se = sd(RPM_normgene)/sqrt(n())) %>% 
    group_by(condition,gene_name) %>% 
    mutate(RPM_cumsum_samplemean = cumsum(RPM_samplemean),
           RPM_cumsum_samplemean_se = cumsum(RPM_samplemean_se),
           RPM_cumsum_samplemean_normgene = cumsum(RPM_samplemean)/sum(RPM_samplemean),
           RPM_cumsum_samplemean_normgene_se = cumsum(RPM_samplemean_se)/sum(RPM_samplemean))
```

# Codon occupancy for each codon - to identify codons with translational defects manifesting as stalling

```{r func_for_codon_occupancy}
# This function takes the average of codon occupancy from each gene

calc_codon_occupancy <- function(data){
    codon_freq <- data %>% ungroup() %>% select(gene_name,codon_seq) %>% 
        group_by(codon_seq) %>% 
        summarise(codon_num = n()) %>% ungroup() %>% 
        mutate(codon_freq = codon_num/sum(codon_num)) 
    
    occupancy_data <- data %>% group_by(gene_name,sample) %>% mutate(codon_num = n()) %>%
        group_by(gene_name,sample,codon_seq) %>% 
        summarise(codon_occupancy_pergene = sum(RPM_normgene)/(n()/mean(codon_num)),codon_count = n()) %>% 
        group_by(sample,codon_seq) %>% 
        summarise(codon_occupancy = mean(codon_occupancy_pergene,na.rm=TRUE))  %>% 
        left_join(.,codon_freq,by="codon_seq") %>% 
        left_join(.,mito_aminoacid_codon,by="codon_seq")
    return(occupancy_data)
}

summarise_codonoccupancy <- function(data){
    data_summary <- data %>% separate(sample,c("condition","replicate"),sep="-") %>% 
        group_by(condition,codon_seq,aminoacid) %>% 
        summarise(avg_codon_occupancy = mean(codon_occupancy),
                  sd_codon_occupancy = sd(codon_occupancy),
                  se_codon_occupancy = sd_codon_occupancy/sqrt(n())) %>% 
      group_by(codon_seq,aminoacid) %>% 
      mutate(occupancy_ratio = avg_codon_occupancy/avg_codon_occupancy[condition == condition1],
             occupancy_ratio_se = occupancy_ratio * 
               sqrt((se_codon_occupancy/avg_codon_occupancy)^2 + 
                      (se_codon_occupancy[condition ==condition1]/avg_codon_occupancy[condition ==condition1])^2)) %>%
      ungroup() %>% mutate(condition = factor(condition,levels = c(condition1,condition2)))
    
    return(data_summary)
}

plot_occupancy <- function(occupancy_data,plot_condition,plot_ratio,add_Met){
  NNG <- c("AAA","AAG","TTA","TTG","CAA","CAG","GAA","GAG","TGA","TGG")
  Met <- c("ATG","ATA")
  occupancy_data <- occupancy_data %>% mutate(plot_type = case_when(!codon_seq %in% c(NNG,Met) ~"A", codon_seq %in% NNG ~ "B", codon_seq %in% Met ~ "C")) %>% unite(aa_codon,c("aminoacid","codon_seq"),sep = "-",remove = FALSE) 
  plot_style <- list(geom_point(), 
                     geom_hline(yintercept = 1),
                     theme(axis.text.x = element_blank(),axis.ticks.x = element_blank(),legend.position = "none"),
                     xlab(""))
  
  # if not plot ratio then we plot cell lines individually with the given input of plot_condition
  if(missing(plot_ratio)){
    output_plot <- occupancy_data %>%  filter(condition == plot_condition) %>% 
      ggplot(aes(x=reorder(codon_seq,avg_codon_occupancy),y = avg_codon_occupancy,col = plot_type)) +
      geom_errorbar(aes(ymin = avg_codon_occupancy - se_codon_occupancy, ymax = avg_codon_occupancy + se_codon_occupancy)) +
      plot_style  + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon)) + 
      ylab("Codon occupancy")
    return(output_plot)
  }
  if(plot_ratio){
    occupancy_data <- occupancy_data  %>%  filter(!condition == condition1) 
    output_plot <- occupancy_data %>% 
      ggplot(aes(x=reorder(codon_seq,occupancy_ratio),y = occupancy_ratio,col = plot_type)) +
      geom_errorbar(aes(ymin = occupancy_ratio - occupancy_ratio_se, ymax = occupancy_ratio + occupancy_ratio_se),col="black") +
      plot_style + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon),col = "black") +
      scale_color_manual(values=c("black","#FF0000","black"))
    if(add_Met){
      output_plot <- output_plot + scale_color_manual(values=c("black","#FF0000","blue")) + 
        geom_text_repel(data = occupancy_data %>% filter(plot_type=="C"),aes(label=aa_codon),col = "black")
    }
    return(output_plot) 
  }
}



```


```{r calc_codon_occupancy}
mito_occupancy <- mito_data %>% calc_codon_occupancy(.)
mito_occupancy_summary <- mito_occupancy %>% summarise_codonoccupancy(.)

mito_occupancy_ratio_summary_plot <- mito_occupancy_summary %>% plot_occupancy(.,plot_ratio = T,add_Met = T) + 
      ylab("Ratio of codon occupancy (MTO1/WT)") + scale_y_continuous(limits = c(0,10))
print(mito_occupancy_ratio_summary_plot)

save_plot("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/TO1_Patient_Plots/Mito_occupancy_ratio_summary_plot.pdf",mito_occupancy_ratio_summary_plot,base_width = 10,base_height=7)

```

# Cumulative sum of codon counts for each gene - to visually identify effects of loss of tRNA modification and stalling sites

```{r func_plot_cumsum}
plot_cumsum <- function(cumsum_data,gene_name,w_facet_wrap){
  plot_style = list( geom_line(),
                     geom_ribbon(aes(linetype=NA, fill=condition, 
                                     ymin =RPM_cumsum_samplemean - RPM_cumsum_samplemean_se, 
                                     ymax = RPM_cumsum_samplemean + RPM_cumsum_samplemean_se),alpha=0.3),
                     scale_y_continuous(labels = function(n) format(n,digits=2,scientific=T)),
                     scale_color_manual(values = mito_pallete),
                     scale_fill_manual(values = mito_pallete),
                     theme(strip.background = element_blank(), aspect.ratio = 0.8)
                     )
  if(missing(w_facet_wrap)){
    mitogenes_cumsum_plot <- cumsum_data  %>% ggplot(aes(x=codon_index,y=RPM_cumsum_samplemean,col=condition)) + 
    plot_style + ggtitle(gene_name)
    return(mitogenes_cumsum_plot)
  }
  if(w_facet_wrap){
     mitogenes_cumsum_plot <- cumsum_data  %>% ggplot(aes(x=codon_index,y=RPM_cumsum_samplemean,col=condition)) + 
    plot_style + facet_wrap(~gene_name,scales = "free") 
  }
  
  return(mitogenes_cumsum_plot)
}
```


```{r plot_cumsum}

mito_data_samplemean_cumsum_plot_all_facet <- mito_data_samplemean %>% plot_cumsum(cumsum_data = .,w_facet_wrap = T)
print(mito_data_samplemean_cumsum_plot_all_facet)

save_plot(plot = mito_data_samplemean_cumsum_plot_all_facet, filename = "D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/TO1_Patient_Plots/Cumsum_plot_samplemean_all_facet.pdf",base_height = 10, base_width =12)

```





# Analysis of MTO1 patient samples
# Load and normalize data

```{r load_data}

raw_data <- read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/HCT116_WT_TO1_PatFib_all_codon_count.txt")


# Load the mapping where we can turn IDs into names that are easily identifiable 
mito_info <-read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/Mitochondrial_Genome_Transcript_GeneInformation.txt") 


condition1 <- "ContFib_Fib"
condition2 <- "Fib_MTO1"
condition_to_analyze <- c(condition1,condition2)

raw_data <- raw_data %>% 
    left_join(.,mito_info,by = c("gene_id"="UCSC_id")) %>% select(gene_name = AssociatedGeneName,GeneType,sample,codon_seq:position_3_count,Complex)  %>% 
  separate(sample, into= c("condition","replicate"),remove = FALSE,sep ="-") %>% 
  filter(condition %in% condition_to_analyze)

# Load the mitochondria codon table 
mito_aminoacid_codon <- read_csv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/AA_Codon_HumMito.csv")

mito_pallete = c("black","red")




```

```{r normalizing_counts}

# Normalize counts in each sample to RPM
mito_data <- raw_data %>% select(-contains("position")) %>% group_by(sample,condition,replicate) %>%  
  mutate(RPM = codon_count_sum/sum(codon_count_sum) * 10^6 + 1) %>% 
  group_by(sample,gene_name) %>%  
  mutate(RPM_normgene = RPM/sum(RPM),
         RPM_cumsum = cumsum(RPM),
         RPM_cumsum_normgene = RPM_cumsum/sum(RPM))  %>% 
  mutate(condition = factor(condition,levels=c(condition1, condition2))) %>% 
  mutate_at(vars(codon_seq),funs(toupper))

# Summarise the replicates and generate cumsum. 
mito_data_samplemean <- mito_data %>% group_by(condition,codon_index,gene_name,codon_seq) %>% 
    summarise(RPM_samplemean = mean(RPM),
              RPM_samplemean_se = sd(RPM)/sqrt(n()),
              RPM_normgene_samplemean = mean(RPM_normgene),
              RPM_normgene_samplemean_se = sd(RPM_normgene)/sqrt(n())) %>% 
    group_by(condition,gene_name) %>% 
    mutate(RPM_cumsum_samplemean = cumsum(RPM_samplemean),
           RPM_cumsum_samplemean_se = cumsum(RPM_samplemean_se),
           RPM_cumsum_samplemean_normgene = cumsum(RPM_samplemean)/sum(RPM_samplemean),
           RPM_cumsum_samplemean_normgene_se = cumsum(RPM_samplemean_se)/sum(RPM_samplemean))
```

# Codon occupancy for each codon - to identify codons with translational defects manifesting as stalling

```{r func_for_codon_occupancy}
# This function takes the average of codon occupancy from each gene

calc_codon_occupancy <- function(data){
    codon_freq <- data %>% ungroup() %>% select(gene_name,codon_seq) %>% 
        group_by(codon_seq) %>% 
        summarise(codon_num = n()) %>% ungroup() %>% 
        mutate(codon_freq = codon_num/sum(codon_num)) 
    
    occupancy_data <- data %>% group_by(gene_name,sample) %>% mutate(codon_num = n()) %>%
        group_by(gene_name,sample,codon_seq) %>% 
        summarise(codon_occupancy_pergene = sum(RPM_normgene)/(n()/mean(codon_num)),codon_count = n()) %>% 
        group_by(sample,codon_seq) %>% 
        summarise(codon_occupancy = mean(codon_occupancy_pergene,na.rm=TRUE))  %>% 
        left_join(.,codon_freq,by="codon_seq") %>% 
        left_join(.,mito_aminoacid_codon,by="codon_seq")
    return(occupancy_data)
}

summarise_codonoccupancy <- function(data){
    data_summary <- data %>% separate(sample,c("condition","replicate"),sep="-") %>% 
        group_by(condition,codon_seq,aminoacid) %>% 
        summarise(avg_codon_occupancy = mean(codon_occupancy),
                  sd_codon_occupancy = sd(codon_occupancy),
                  se_codon_occupancy = sd_codon_occupancy/sqrt(n())) %>% 
      group_by(codon_seq,aminoacid) %>% 
      mutate(occupancy_ratio = avg_codon_occupancy/avg_codon_occupancy[condition == condition1],
             occupancy_ratio_se = occupancy_ratio * 
               sqrt((se_codon_occupancy/avg_codon_occupancy)^2 + 
                      (se_codon_occupancy[condition ==condition1]/avg_codon_occupancy[condition ==condition1])^2)) %>%
      ungroup() %>% mutate(condition = factor(condition,levels = c(condition1,condition2)))
    
    return(data_summary)
}

plot_occupancy <- function(occupancy_data,plot_condition,plot_ratio,add_Met){
  NNG <- c("AAA","AAG","TTA","TTG","CAA","CAG","GAA","GAG","TGA","TGG")
  Met <- c("ATG","ATA")
  occupancy_data <- occupancy_data %>% mutate(plot_type = case_when(!codon_seq %in% c(NNG,Met) ~"A", codon_seq %in% NNG ~ "B", codon_seq %in% Met ~ "C")) %>% unite(aa_codon,c("aminoacid","codon_seq"),sep = "-",remove = FALSE) 
  plot_style <- list(geom_point(), 
                     geom_hline(yintercept = 1),
                     theme(axis.text.x = element_blank(),axis.ticks.x = element_blank(),legend.position = "none"),
                     xlab(""))
  
  # if not plot ratio then we plot cell lines individually with the given input of plot_condition
  if(missing(plot_ratio)){
    output_plot <- occupancy_data %>%  filter(condition == plot_condition) %>% 
      ggplot(aes(x=reorder(codon_seq,avg_codon_occupancy),y = avg_codon_occupancy,col = plot_type)) +
      geom_errorbar(aes(ymin = avg_codon_occupancy - se_codon_occupancy, ymax = avg_codon_occupancy + se_codon_occupancy)) +
      plot_style  + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon)) + 
      ylab("Codon occupancy")
    return(output_plot)
  }
  if(plot_ratio){
    occupancy_data <- occupancy_data  %>%  filter(!condition == condition1) 
    output_plot <- occupancy_data %>% 
      ggplot(aes(x=reorder(codon_seq,occupancy_ratio),y = occupancy_ratio,col = plot_type)) +
      geom_errorbar(aes(ymin = occupancy_ratio - occupancy_ratio_se, ymax = occupancy_ratio + occupancy_ratio_se),col="black") +
      plot_style + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon),col = "black") +
      scale_color_manual(values=c("black","#FF0000","black"))
    if(add_Met){
      output_plot <- output_plot + scale_color_manual(values=c("black","#FF0000","blue")) + 
        geom_text_repel(data = occupancy_data %>% filter(plot_type=="C"),aes(label=aa_codon),col = "black")
    }
    return(output_plot) 
  }
}


plot_indiv_occupancy <- function(occupancy_data,plot_condition,plot_ratio){
  NNG <- c("AAA","AAG","TTA","TTG","CAA","CAG","GAA","GAG","TGA","TGG")
  occupancy_data <- occupancy_data %>% mutate(plot_type = case_when(!codon_seq %in% c(NNG) ~"A", codon_seq %in% NNG ~ "B"))  %>% 
  separate(sample, into= c("condition","replicate"),remove = FALSE,sep ="-") %>% group_by(codon_seq) %>% mutate(occupancy_ratio = codon_occupancy/mean(codon_occupancy[condition == plot_condition[2]]))
  plot_style <- list(geom_point(), 
                     geom_hline(yintercept = 1),
                     theme(axis.text.x = element_blank(),axis.ticks.x = element_blank(),legend.position = "none"),
                     xlab(""))
  
  # if not plot ratio then we plot cell lines individually with the given input of plot_condition
  if(missing(plot_ratio)){
    output_plot <- occupancy_data %>%  filter(sample == plot_condition[1]) %>% 
      ggplot(aes(x=reorder(codon_seq,avg_codon_occupancy),y = avg_codon_occupancy,col = plot_type)) +
      geom_errorbar(aes(ymin = avg_codon_occupancy - se_codon_occupancy, ymax = avg_codon_occupancy + se_codon_occupancy)) +
      plot_style  + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon)) + 
      ylab("Codon occupancy")
    return(output_plot)
  }
  if(plot_ratio){
    occupancy_data <- occupancy_data  %>%  filter(sample == plot_condition[1]) 
    output_plot <- occupancy_data %>% 
      ggplot(aes(x=reorder(codon_seq,occupancy_ratio),y = occupancy_ratio,col = plot_type)) +
      plot_style + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=codon_seq),col = "black") +
      scale_color_manual(values=c("black","#FF0000","black"))
    return(output_plot) 
  }
}

```


```{r calc_codon_occupancy}
mito_occupancy <- mito_data %>% calc_codon_occupancy(.)
mito_occupancy_summary <- mito_occupancy %>% summarise_codonoccupancy(.)

mito_occupancy_ratio_summary_plot <- mito_occupancy_summary %>% plot_occupancy(.,plot_ratio = T,add_Met = T) + 
      ylab("Ratio of codon occupancy (MTO1_pat/WT)") + scale_y_continuous(limits = c(0,12), breaks=seq(0,12,2))
print(mito_occupancy_ratio_summary_plot)

save_plot("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/TO1_Patient_Plots/MTO1_pat_Mito_occupancy_ratio_summary_plot.pdf",mito_occupancy_ratio_summary_plot,base_width = 10,base_height=7)


mito_occupancy  %>% plot_indiv_occupancy(.,plot_condition=c("Fib_MTO1-1","ContFib_Fib"),plot_ratio = T) + 
      ylab("Ratio of codon occupancy (MTO1-1_pat/WT)") + scale_y_continuous(limits = c(0,12), breaks=seq(0,12,2))
mito_occupancy  %>% plot_indiv_occupancy(.,plot_condition=c("Fib_MTO1-2","ContFib_Fib"),plot_ratio = T) + 
      ylab("Ratio of codon occupancy (MTO1-2_pat/WT)") + scale_y_continuous(limits = c(0,12), breaks=seq(0,12,2))
```



# Analysis of MELAS patient samples
# Load and normalize data

```{r load_data}

raw_data <- read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/HCT116_WT_TO1_PatFib_all_codon_count.txt")


# Load the mapping where we can turn IDs into names that are easily identifiable 
mito_info <-read_tsv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/Mitochondrial_Genome_Transcript_GeneInformation.txt") 


condition1 <- "ContFib_Fib"
condition2 <- "Fib_MELAS"
condition_to_analyze <- c(condition1,condition2)

raw_data <- raw_data %>% 
    left_join(.,mito_info,by = c("gene_id"="UCSC_id")) %>% select(gene_name = AssociatedGeneName,GeneType,sample,codon_seq:position_3_count,Complex)  %>% 
  separate(sample, into= c("condition","replicate"),remove = FALSE,sep ="-") %>% 
  filter(condition %in% condition_to_analyze)

# Load the mitochondria codon table 
mito_aminoacid_codon <- read_csv("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/AA_Codon_HumMito.csv")

mito_pallete = c("black","red")




```

```{r normalizing_counts}

# Normalize counts in each sample to RPM
mito_data <- raw_data %>% select(-contains("position")) %>% group_by(sample,condition,replicate) %>%  
  mutate(RPM = codon_count_sum/sum(codon_count_sum) * 10^6 + 1) %>% 
  group_by(sample,gene_name) %>%  
  mutate(RPM_normgene = RPM/sum(RPM),
         RPM_cumsum = cumsum(RPM),
         RPM_cumsum_normgene = RPM_cumsum/sum(RPM))  %>% 
  mutate(condition = factor(condition,levels=c(condition1, condition2))) %>% 
  mutate_at(vars(codon_seq),funs(toupper))

# Summarise the replicates and generate cumsum. 
mito_data_samplemean <- mito_data %>% group_by(condition,codon_index,gene_name,codon_seq) %>% 
    summarise(RPM_samplemean = mean(RPM),
              RPM_samplemean_se = sd(RPM)/sqrt(n()),
              RPM_normgene_samplemean = mean(RPM_normgene),
              RPM_normgene_samplemean_se = sd(RPM_normgene)/sqrt(n())) %>% 
    group_by(condition,gene_name) %>% 
    mutate(RPM_cumsum_samplemean = cumsum(RPM_samplemean),
           RPM_cumsum_samplemean_se = cumsum(RPM_samplemean_se),
           RPM_cumsum_samplemean_normgene = cumsum(RPM_samplemean)/sum(RPM_samplemean),
           RPM_cumsum_samplemean_normgene_se = cumsum(RPM_samplemean_se)/sum(RPM_samplemean))
```

# Codon occupancy for each codon - to identify codons with translational defects manifesting as stalling

```{r func_for_codon_occupancy}
# This function takes the average of codon occupancy from each gene

calc_codon_occupancy <- function(data){
    codon_freq <- data %>% ungroup() %>% select(gene_name,codon_seq) %>% 
        group_by(codon_seq) %>% 
        summarise(codon_num = n()) %>% ungroup() %>% 
        mutate(codon_freq = codon_num/sum(codon_num)) 
    
    occupancy_data <- data %>% group_by(gene_name,sample) %>% mutate(codon_num = n()) %>%
        group_by(gene_name,sample,codon_seq) %>% 
        summarise(codon_occupancy_pergene = sum(RPM_normgene)/(n()/mean(codon_num)),codon_count = n()) %>% 
        group_by(sample,codon_seq) %>% 
        summarise(codon_occupancy = mean(codon_occupancy_pergene,na.rm=TRUE))  %>% 
        left_join(.,codon_freq,by="codon_seq") %>% 
        left_join(.,mito_aminoacid_codon,by="codon_seq")
    return(occupancy_data)
}

summarise_codonoccupancy <- function(data){
    data_summary <- data %>% separate(sample,c("condition","replicate"),sep="-") %>% 
        group_by(condition,codon_seq,aminoacid) %>% 
        summarise(avg_codon_occupancy = mean(codon_occupancy),
                  sd_codon_occupancy = sd(codon_occupancy),
                  se_codon_occupancy = sd_codon_occupancy/sqrt(n())) %>% 
      group_by(codon_seq,aminoacid) %>% 
      mutate(occupancy_ratio = avg_codon_occupancy/avg_codon_occupancy[condition == condition1],
             occupancy_ratio_se = occupancy_ratio * 
               sqrt((se_codon_occupancy/avg_codon_occupancy)^2 + 
                      (se_codon_occupancy[condition ==condition1]/avg_codon_occupancy[condition ==condition1])^2)) %>%
      ungroup() %>% mutate(condition = factor(condition,levels = c(condition1,condition2)))
    
    return(data_summary)
}

plot_occupancy <- function(occupancy_data,plot_condition,plot_ratio,add_Met){
  NNG <- c("AAA","AAG","TTA","TTG","CAA","CAG","GAA","GAG","TGA","TGG")
  Met <- c("ATG","ATA")
  occupancy_data <- occupancy_data %>% mutate(plot_type = case_when(!codon_seq %in% c(NNG,Met) ~"A", codon_seq %in% NNG ~ "B", codon_seq %in% Met ~ "C")) %>% unite(aa_codon,c("aminoacid","codon_seq"),sep = "-",remove = FALSE) 
  plot_style <- list(geom_point(), 
                     geom_hline(yintercept = 1),
                     theme(axis.text.x = element_blank(),axis.ticks.x = element_blank(),legend.position = "none"),
                     xlab(""))
  
  # if not plot ratio then we plot cell lines individually with the given input of plot_condition
  if(missing(plot_ratio)){
    output_plot <- occupancy_data %>%  filter(condition == plot_condition) %>% 
      ggplot(aes(x=reorder(codon_seq,avg_codon_occupancy),y = avg_codon_occupancy,col = plot_type)) +
      geom_errorbar(aes(ymin = avg_codon_occupancy - se_codon_occupancy, ymax = avg_codon_occupancy + se_codon_occupancy)) +
      plot_style  + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon)) + 
      ylab("Codon occupancy")
    return(output_plot)
  }
  if(plot_ratio){
    occupancy_data <- occupancy_data  %>%  filter(!condition == condition1) 
    output_plot <- occupancy_data %>% 
      ggplot(aes(x=reorder(codon_seq,occupancy_ratio),y = occupancy_ratio,col = plot_type)) +
      geom_errorbar(aes(ymin = occupancy_ratio - occupancy_ratio_se, ymax = occupancy_ratio + occupancy_ratio_se),col="black") +
      plot_style + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon),col = "black") +
      scale_color_manual(values=c("black","#FF0000","black"))
    if(add_Met){
      output_plot <- output_plot + scale_color_manual(values=c("black","#FF0000","blue")) + 
        geom_text_repel(data = occupancy_data %>% filter(plot_type=="C"),aes(label=aa_codon),col = "black")
    }
    return(output_plot) 
  }
}



plot_indiv_occupancy <- function(occupancy_data,plot_condition,plot_ratio){
  NNG <- c("AAA","AAG","TTA","TTG","CAA","CAG","GAA","GAG","TGA","TGG")
  occupancy_data <- occupancy_data %>% mutate(plot_type = case_when(!codon_seq %in% c(NNG) ~"A", codon_seq %in% NNG ~ "B"))  %>% 
  separate(sample, into= c("condition","replicate"),remove = FALSE,sep ="-") %>% group_by(codon_seq) %>% mutate(occupancy_ratio = codon_occupancy/mean(codon_occupancy[condition == plot_condition[2]]))
  plot_style <- list(geom_point(), 
                     geom_hline(yintercept = 1),
                     theme(axis.text.x = element_blank(),axis.ticks.x = element_blank(),legend.position = "none"),
                     xlab(""))
  
  # if not plot ratio then we plot cell lines individually with the given input of plot_condition
  if(missing(plot_ratio)){
    output_plot <- occupancy_data %>%  filter(sample == plot_condition[1]) %>% 
      ggplot(aes(x=reorder(codon_seq,avg_codon_occupancy),y = avg_codon_occupancy,col = plot_type)) +
      geom_errorbar(aes(ymin = avg_codon_occupancy - se_codon_occupancy, ymax = avg_codon_occupancy + se_codon_occupancy)) +
      plot_style  + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=aa_codon)) + 
      ylab("Codon occupancy")
    return(output_plot)
  }
  if(plot_ratio){
    occupancy_data <- occupancy_data  %>%  filter(sample == plot_condition[1]) 
    output_plot <- occupancy_data %>% 
      ggplot(aes(x=reorder(codon_seq,occupancy_ratio),y = occupancy_ratio,col = plot_type)) +
      plot_style + geom_text_repel(data = occupancy_data %>% filter(plot_type== "B"),aes(label=codon_seq),col = "black") +
      scale_color_manual(values=c("black","#FF0000","black"))
    return(output_plot) 
  }
}
```


```{r calc_codon_occupancy}
mito_occupancy <- mito_data %>% calc_codon_occupancy(.)
mito_occupancy_summary <- mito_occupancy %>% summarise_codonoccupancy(.)

mito_occupancy_ratio_summary_plot <- mito_occupancy_summary %>% plot_occupancy(.,plot_ratio = T,add_Met = T) + 
      ylab("Ratio of codon occupancy (MELAS_pat/WT)") + scale_y_continuous(limits = c(0,6), breaks=seq(0,6,1))
print(mito_occupancy_ratio_summary_plot)

save_plot("D:/Sequencing_Data_1CTRNA/RiboSeqAnalysis_Resubmission/TO1_Patient_Plots/MELAS_pat_Mito_occupancy_ratio_summary_plot.pdf",mito_occupancy_ratio_summary_plot,base_width = 10,base_height=7)


mito_occupancy  %>% plot_indiv_occupancy(.,plot_condition=c("Fib_MELAS-1","ContFib_Fib"),plot_ratio = T) + 
      ylab("Ratio of codon occupancy (MELAS-1_pat/WT)") + scale_y_continuous(limits = c(0,6), breaks=seq(0,6,1))
mito_occupancy  %>% plot_indiv_occupancy(.,plot_condition=c("Fib_MELAS-2","ContFib_Fib"),plot_ratio = T) + 
      ylab("Ratio of codon occupancy (MELAS-2_pat/WT)") + scale_y_continuous(limits = c(0,6), breaks=seq(0,6,1))

```