---
title: "Load Data"
author: "Lance Parsons <lparsons@princeton.edu>"
date: "July 20, 2016"
output: html_document
---

## Load all sample data into DESeq

First we'll setup some variable to keep track of the input and output directories.

```{r directory_setup}
# Input and output directories
data_dir <- "data"
output_dir <- "output"
dir.create(output_dir, showWarnings = FALSE)
```

We'll use a table we created in Excel to keep track of the samples, the files that contain the data for those samples, and any relevant experimental factors (*e.g.* treatment, batch, cell line, etc.)

```{r sample_table}
# Load sample table from csv (Excel) file
sampleTable <- read.csv("sample_table.csv")
sampleTable
```

Finally, we'll use that table to actually load the data into R using DESeq2.

```{r load_data}
# Load count data into DESeq2
ddsFull <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable, 
                                      directory = data_dir, 
                                      design = ~1)
ddsFull
```
