---
title: "`r params$contrast_factor`: `r params$numerator` vs. `r params$denominator`"
author: "Lance Parsons <lparsons@princeton.edu>"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
params:
    contrast_factor: "sample"
    numerator: "S11"
    denominator: "WTC"
---

```{r child='init_session.Rmd'}
```

```{r child='load_data.Rmd'}
```


## Select samples


```{r select_samples}
# Select samples where `batch` is `2015-10`
dds <- ddsFull[, ddsFull$batch == "2015-10" & 
                   ddsFull@colData[,params$contrast_factor] %in%
                                        c(params$numerator, params$denominator)]
```

Sometimes it is necessary to drop levels of the factors, in case that all the
samples for one or more levels of a factor in the design have been removed. If
time were included in the design formula, the following code could be used to
take care of dropped levels in this column.
```{r droplevels}
# Uncomment these lines if you selected a subset of samples above
dds$colony <- droplevels( dds$colony )
dds$genotype <- droplevels( dds$genotype )
dds$sample <- droplevels( dds$sample )
```

## Setup Experimental Design

### Setup design
The experimental `design` defines the factors (attributes) of the model and how 
they affect the transcription level.

The `contrast` determines what factor and which values of that factor we
want to compare to get a differntial epxression.
```{r setup_design}
# Experimental Design
design(dds) = ~ sample

# Contrast
contrast <- c(params$contrast_factor, params$numerator, params$denominator)

# Set base name for output files
output_basename <- sprintf("2015-10_%s_%s_vs_%s_standard_analysis", 
                           params$contrast_factor, 
                           params$numerator, 
                           params$denominator)
```


```{r child='deseq2_standard_analysis_core.Rmd'}
```

```{r child='annotate_results_orgdb.Rmd'}
```

## Display the version of R as well as any loaded packages.
```{r sessionInfo}
sessionInfo()
```
